---
title: "Divvy Shared Bike 2018 data preparation"
author: "Pornsinee Pongpetrarat"
date: "1/20/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### About the data
This dataset is provided by Divvy Shared Bikes company from Chicago and their stations geographic coordinate are from Chicago Data Portal

### Data Preparation

We had 2 datasets:

1. 2018 Divvy Dataset : this dataset has everything except for geographic coordinate, which is important to create map in Tableau later on
2. Divvy stations geographic coordinate dataset : this dataset has stations’ geographic coordinate, we will join this dataset to Divvy Dataset

install necessary packages then load library
```{r}
library(tidyverse)
library(geosphere)
```


Import 2018 Divvy dataset and Divvy stations geographic coordinate dataset
```{r}
 temp = list.files(pattern="Divvy_Trips_2018.*.csv", recursive = TRUE)
 Divvy_12_months <- lapply(temp, read.csv) %>%
  bind_rows()
 
 Station_position <- read.csv("Divvy_Bicycle_Stations_-_In_Service.csv") 
```


check for duplicate

if sum = 0 then no duplicate. This is because in logical data type, TRUE = 1 and FALSE = 0
```{r}
duplicated(Divvy_12_months$trip_id)%>%
  sum()
duplicated(Station_position$Station.Name)%>%
  sum()
```


check for NA

if sum = 0 then no NA
```{r}
duplicated(Divvy_12_months$trip_id)%>%
  sum()
duplicated(Station_position$Station.Name)%>%
  sum()
```


add age column
```{r}
Divvy_12_months <- Divvy_12_months%>%
  mutate(age = 2018 - birthyear)
```


change trip duration data type from chr–>num and add trip duration(minute) column
```{r}
Divvy_12_months$tripduration <- gsub(",", "", Divvy_12_months$tripduration)
Divvy_12_months$tripduration <- as.numeric(Divvy_12_months$tripduration)
Divvy_12_months <- Divvy_12_months%>%
  mutate(tripduration_min = round(tripduration/60))
```


add latitude, longitude

In this part, I select columns from Divvy dataset to create new lists so I cฟื do trial and error as many times as I want without affecting Divvy_12_month, so I didn’t have to reload Divvy dataset everytime I messed up.

first, create 3 new lists

1. 2 lists contain station name and station id of start stations(from station) while the other of end stations(to station) from Divvy_12_months
2. 1 list contains station id, latitude, longitude from Station_position
note: from station = start station and to station = end station

create 3 lists
```{r}
select_from_station_name <- Divvy_12_months %>%
  select(from_station_name,from_station_id)
select_to_station_name <- Divvy_12_months %>%
  select(to_station_name, to_station_id)
select_station_position <- Station_position%>%
  select(ID, Latitude, Longitude)
```


left join stations geo coordinate to start station and end station
```{r}
select_from_station_name <- select_from_station_name%>%
  left_join(select_station_position , by= c('from_station_id' = 'ID'))
select_to_station_name <- select_to_station_name%>%
  left_join(select_station_position , by= c('to_station_id' = 'ID'))
```


check for NA
```{r}
lapply(lapply(select_from_station_name, is.na), sum)
lapply(lapply(select_to_station_name, is.na), sum)
```
there are 146,795 of NA rows in latitude and longitude columns from start station list

there are 156,632 of NA rows in latitude and longitude columns from end station list


check for blank string in station names columns
```{r}
select_from_station_name%>%
  filter(from_station_name == "")
select_to_station_name%>%
  filter(to_station_name == "")
```


Check for impact if I decide to drop NA rows in latitude or longitude columns

first, combind columns together
```{r}
from_to_station_latlng <- cbind(select_from_station_name,select_to_station_name)
colnames(from_to_station_latlng)[3] <- "from_Latitude"
colnames(from_to_station_latlng)[4] <- "from_Longitude"
colnames(from_to_station_latlng)[7] <- "to_Latitude"
colnames(from_to_station_latlng)[8] <- "to_Longitude"
```


count for all NA rows I have to drop
```{r}
from_to_station_latlng%>%
  filter(is.na(from_Latitude) == TRUE | is.na(from_Longitude) == TRUE |
           is.na(to_Latitude) == TRUE | is.na(to_Longitude) == TRUE) %>%
  count()
```


If we decided to drop rows with NA in latitude or longitude column we will have to drop 293,131 rows of data. 

I decided to drop those NA rows.

combind start and end station latitude, longitude columns to main Divvy_12_months
```{r}
Divvy_12_months <- cbind(Divvy_12_months, from_Latitude = from_to_station_latlng[,3])
Divvy_12_months <- cbind(Divvy_12_months, from_Logitude = from_to_station_latlng[,4])
Divvy_12_months <- cbind(Divvy_12_months, to_Latitude = from_to_station_latlng[,7])
Divvy_12_months <- cbind(Divvy_12_months, to_Logitude = from_to_station_latlng[,8])
```


drop the NA rows
```{r}
Divvy_12_months_no_NA_latlng <- Divvy_12_months[complete.cases(Divvy_12_months[,15:18]),]
```


check if we really delete 293,131 rows
```{r}
nrow(Divvy_12_months)-nrow(Divvy_12_months_no_NA_latlng)
```


check for NA in latitude, longitude columns
```{r}
sum(is.na(Divvy_12_months_no_NA_latlng[,15:18]))
```


check for negative trip duration
```{r}
Divvy_12_months_no_NA_latlng%>%
  filter(tripduration_min < 0)
```


check for more than 24 hour trip
```{r}
Divvy_12_months_no_NA_latlng%>%
  filter(tripduration_min> 24*60)%>%
  count()
```
there were 1301 rows

filter out those 1301 rows
```{r}
Divvy_12_months_clean <- Divvy_12_months_no_NA_latlng%>%
  filter(tripduration_min <= 24*60)
```


write csv file for Tableau
```{r}
Divvy_12_months_clean %>%
  write.csv("Divvy2018_with_latlng.csv")
```



